---
title: 
author: 
output: 
  html_notebook
runtime: shiny
---
## Brain Atrophy in Normal Aging
#### by Lauren N. Koenig
Displayed below is data from the Normal Aging Cohort described in [this publication](https://pubmed.ncbi.nlm.nih.gov/32334404/), and at AAIC 2021. Recalculating everything in R can be slow so **please be patient**. The participants are from DIAN and OASIS databases, are amyloid PET negative, and remained cognitively normal for at least 3 year after the displayed MRI (processed with Freesurfer, primarily version 5.3). The exception are the amyloid+ participants which can be added, who were also cognitively normal at time of MRI but were amyloid PET positive.
This work was supported by the following grants: U19AG03243, P50AG00561, P30NS09857781, P01AG026276, P01AG003991, R01AG043434, UL1TR000448, R01EB009352, P50AG005681, P01AG026276


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```


```{r Load and Clean Up Data, include=FALSE}
#Load the R packages necessary for rest of script
library("cowplot")
library("ggseg")
library("ggplot2")
library("ggrepel")
library("tidyverse")
library("shiny")
library("lm.beta")
library("toOrdinal")


#load data
load("Normal_Aging_Input_Shiny_2021-07-08.RData")

#Set normals data's min and max ages
n_min <- ceiling(min(normals$Age, na.rm=TRUE))
n_max <- floor(max(normals$Age, na.rm=TRUE))

#prevents scientific notation
options(scipen=999)

region_list <- colnames(select(normals, `Banks STS Thickness`:`5th Ventricle Volume`))


#For debugging:
# input <- data.frame(row.names="input")
# input$span=0.7
# input$sex=1
# input$pcad=FALSE
# input$apoe4=1
# input$ROI_name="Hippocampus Volume"
# input$individual_points=TRUE
# input$Age=60
# input$ICV=1500000
# input$region_val=4000

footer = ""
```

```{r make normal aging curve}
ui <- shinyUI(fluidPage(sidebarLayout(
  sidebarPanel(
    h4(strong("Change Parameters:")),
    checkboxInput("pcad",label="Include Amyloid+ (preclinical AD)"),
    radioButtons("sex", label=h5("Sex"), choices=list("All"=1, "Men"=2, "Women"=3)),
    radioButtons("apoe4", label=h5("APOE Status"), choices=list("All"=1, "E4 Non-Carriers"=2, "E4 Carriers"=3)),
    sliderInput("span", label=h5("Alpha (smoothing)"), min=0.1, value=0.7, max=2, step=0.1 ),
    width=2),
  mainPanel(
    tabsetPanel(type="tabs",
      tabPanel("Aging Curves", 
        fluidRow(
          column(3, selectInput("ROI_name", label=h5("Select region"), choices=region_list, selected="Hippocampus Volume")
          ),
          
          column(9,
            fluidRow( 
                     column(2, ""),
                     column(6, h5(strong("Input an individual's data:")))
                     ),
            fluidRow(
              column(2, checkboxInput("individual_points",label="Show all datapoints", value=TRUE)),
              column(2, offset = 0, numericInput("Age", label=h6("Age (years)"), value=NA)),
              column(2, numericInput("region_val", label=h6("Volume/Thickness"), value=NA)), 
              column(3, numericInput("ICV", label=h6("Intracranial Volume"), value=1500000))
            )
          )
        ),
        hr(),
        plotOutput("region_plot",  width = "1100px", height = "500px"),
      ),
      tabPanel("Whole Brain Map",
          paste("The strength of atrophy is measured as the standardized beta weight of a linear model between a region's volume/thickness and age. The pattern of atrophy is similarly measured but between the estimated derivative of the data and age (the lower graph in the Aging Curves tab)."),
          plotOutput("ggseg_plot", width = "850px", height = "800px")),    
      tabPanel("Whole Brain Table",
          paste("The strength of atrophy is measured as the standardized beta weight of a linear model between a region's volume/thickness and age. The pattern of atrophy is similarly measured but between the estimated derivative of the data and age (the lower graph in the Aging Curves tab)."),
          tableOutput("table"))
      ))),
  hr(),
  fluidRow(footer)
  ))

  
server <- shinyServer(function(input, output) {  
  
  #define function to calculation data
  fit_data <- function(normals, region, span_input){
        smoothFit <- loess( as.matrix(normals[,region]) ~ normals$Age, degree=1, span=span_input) #locally fits a polynomial surface
        smoothPred <- predict( smoothFit, newdata=n_min:n_max, se=TRUE ) #uses loess fit to predict data for each age
        stand_dev2 <- matrix()
        for (i in 1:length(n_min:n_max)) {
          stand_dev2[i] <- sd(as.matrix(normals[normals$Age > seq(n_min,n_max, by=1)[i]-1 & normals$Age < seq(n_min,n_max, by=1)[i]+1,][,region]))
        }
        sd_fit <- loess(stand_dev2 ~ c(n_min:n_max), degree=1, span=span_input)
        sd_pred <- predict(sd_fit, newdata=n_min:n_max)
        plotFrame <- data.frame( Age=c(n_min:n_max), fit=smoothPred$fit, sd=sd_pred ) #combines all the model's data into single dataframe

        return(plotFrame)
  }
  
  region_map_data <- function(normals){
    mc <- "holm"

    #make dataframes to hold results
    results <- data.frame() #results from normal aging cohort
    slopes <- data.frame(Age=(n_min+1):n_max)
    
    age_range <- 1 #range at which slopes are calculated (1=1 year bin)
      
    for (region in region_list){
      ####linear pattern
      model <- lm.beta(lm(normals[,region] ~ normals$Age ))
      results[region, "strength_p"] <- summary(model)$coefficients[2,5]
      results[region, "strength_beta"] <- summary(model)$coefficients[2,2]
  
      ####non-linear pattern
      plotFrame <- fit_data(normals, region, input$span)

      slope <- diff(plotFrame$fit, differences=1)/age_range #normalize to range fit the data (if using an age increase other than 1 year it switches it to per year)
      slope <- (slope) / plotFrame$fit[-length(plotFrame$fit)] #make it percent change

      #save data to dataframe
      slopes[,region] <- slope

      #make lm for each region and each variable in normals list, save p-value
       model <- lm.beta(lm(slopes[,region] ~ slopes$Age ))
       results[region, "pattern_p"] <- summary(model)$coefficients[2,5]
       results[region, "pattern_beta"] <- summary(model)$coefficients[2,2]
    }  
    
  #Once done finding all p-values, do multiple comparisons correction
  results[,"strength_p"] <- p.adjust((results[,"strength_p"]), method=mc)
  results[,"pattern_p"] <- p.adjust((results[,"pattern_p"]), method=mc)

   for (i in c("strength_beta", "pattern_beta")){results[i] <- round(results[i], digits=3)}
    
    for (i in c("strength_p", "pattern_p")){
      for (j in 1:length(results[,i])) {
        if (results[j,i] >=0.01) {results[j,i] <- format(round(as.numeric(results[j,i]), digits=2), nsmall=2)}
        else if (results[j,i]<0.01 & results[j,i]>=0.001 ){results[j,i] <- round(as.numeric(results[j,i]), digits=3)}
        else if (results[j,i]<0.001){results[j,i] <- "<0.001" }
      }
    }
  results$Region <- rownames(results)
  results <- relocate(results, Region)
  
 

  
        return(results)
  } 
  
  ggseg_graph <- function(results){
     
        
        
        #reverse values for making images
      reverse <- c("Lateral Ventricles Volume", "Choroid Plexus Volume" , "CSF Volume", "Inferior Lateral Ventricles Volume", "White Matter Hyperintensities Volume", "Non-WM Hyperintensities Volume", "Optic Chiasm Volume", "Vessel Volume", "3rd Ventricle Volume", "4th Ventricle Volume", "5th Ventricle Volume"  )
      
      
      #flip linear model beta values where it makes sense
      results_adjusted <- results
      results_adjusted[ row.names(results_adjusted) %in% reverse,]$strength_beta <- -1*results_adjusted[ row.names(results_adjusted) %in% reverse,]$strength_beta
      results_adjusted[ row.names(results_adjusted) %in% reverse,]$pattern_beta <- -1*results_adjusted[ row.names(results_adjusted) %in% reverse,]$pattern_beta
    
      #select out thickness measures and rename to match ggseg conventions
      thicknesses <- results_adjusted[c("Banks STS Thickness", "Caudal Anterior Cingulate Thickness", "Caudal Middle Frontal Thickness", "Cuneus Thickness", "Entorhinal Thickness", "Frontal Pole Thickness", "Fusiform Thickness", "Inferior Parietal Thickness", "Inferior Temporal Thickness", "Insula Thickness",  "Isthmus Cingulate Thickness", "Lateral Occipital Thickness","Lateral Orbitofrontal Thickness", "Lingual Thickness", "Medial Orbitofrontal Thickness", "Middle Temporal Thickness", "Paracentral Thickness", "Parahippocampal Thickness", "Pars Opercularis Thickness", "Pars Orbitalis Thickness", "Pars Triangularis Thickness", "Pericalcarine Thickness", "Postcentral Thickness", "Posterior Cingulate Thickness", "Precentral Thickness", "Precuneus Thickness", "Rostral Anterior Cingulate Thickness", "Rostral Middle Frontal Thickness", "Superior Frontal Thickness", "Superior Parietal Thickness", "Superior Temporal Thickness", "Supramarginal Thickness", "Temporal Pole Thickness", "Transverse Temporal Thickness"),]

      thicknesses$Region <- dplyr::recode(thicknesses$Region, "Banks STS Thickness" = "banks superior temporal", "Caudal Anterior Cingulate Thickness" = "caudal anterior cingulate", "Caudal Middle Frontal Thickness" = "caudal middle frontal", "Cuneus Thickness" = "cuneus", "Entorhinal Thickness" = "entorhinal", "Frontal Pole Thickness" = "frontal pole", "Fusiform Thickness" = "fusiform", "Inferior Parietal Thickness" = "inferior parietal", "Inferior Temporal Thickness" = "inferior temporal",  "Insula Thickness" = "insula", "Isthmus Cingulate Thickness" = "isthmus cingulate", "Lateral Occipital Thickness" = "lateral occipital",  "Lateral Orbitofrontal Thickness" = "lateral orbitofrontal", "Lingual Thickness" = "lingual", "Medial Orbitofrontal Thickness" = "medial orbito frontal", "Middle Temporal Thickness" = "middle temporal", "Paracentral Thickness" = "para central", "Parahippocampal Thickness" = "parahippocampal", "Pars Opercularis Thickness" = "pars opercularis", "Pars Orbitalis Thickness" = "pars orbitalis", "Pars Triangularis Thickness" = "pars triangularis", "Pericalcarine Thickness" = "pericalcarine", "Postcentral Thickness" = "post central", "Posterior Cingulate Thickness" = "posterior cingulate", "Precentral Thickness" = "pre central", "Precuneus Thickness" = "precuneus", "Rostral Anterior Cingulate Thickness" = "rostral anterior cingulate", "Rostral Middle Frontal Thickness" = "rostral middle frontal", "Superior Frontal Thickness" = "superior frontal", "Superior Parietal Thickness" = "superior parietal", "Superior Temporal Thickness" = "superior temporal", "Supramarginal Thickness" = "supramarginal", "Temporal Pole Thickness" = "temporal pole", "Transverse Temporal Thickness" = "transverse temporal")
      rownames(thicknesses) <- thicknesses$Region
      
      #select out volume measures and rename to match ggseg conventions
      volumes <- results_adjusted[c("Banks STS Volume", "Caudal Anterior Cingulate Volume", "Caudal Middle Frontal Volume", "Cuneus Volume", "Entorhinal Volume", "Frontal Pole Volume", "Fusiform Volume", "Inferior Parietal Volume", "Inferior Temporal Volume", "Insula Volume",  "Isthmus Cingulate Volume", "Lateral Occipital Volume","Lateral Orbitofrontal Volume", "Lingual Volume", "Medial Orbitofrontal Volume", "Middle Temporal Volume", "Paracentral Volume", "Parahippocampal Volume", "Pars Opercularis Volume", "Pars Orbitalis Volume", "Pars Triangularis Volume", "Pericalcarine Volume", "Postcentral Volume", "Posterior Cingulate Volume", "Precentral Volume", "Precuneus Volume", "Rostral Anterior Cingulate Volume", "Rostral Middle Frontal Volume", "Superior Frontal Volume", "Superior Parietal Volume", "Superior Temporal Volume", "Supramarginal Volume", "Temporal Pole Volume", "Transverse Temporal Volume"),]
      
      volumes$Region <- dplyr::recode(volumes$Region, "Banks STS Volume" = "banks superior temporal", "Caudal Anterior Cingulate Volume" = "caudal anterior cingulate", "Caudal Middle Frontal Volume" = "caudal middle frontal", "Cuneus Volume" = "cuneus", "Entorhinal Volume" = "entorhinal", "Frontal Pole Volume" = "frontal pole", "Fusiform Volume" = "fusiform", "Inferior Parietal Volume" = "inferior parietal", "Inferior Temporal Volume" = "inferior temporal",  "Insula Volume" = "insula", "Isthmus Cingulate Volume" = "isthmus cingulate", "Lateral Occipital Volume" = "lateral occipital",  "Lateral Orbitofrontal Volume" = "lateral orbitofrontal", "Lingual Volume" = "lingual", "Medial Orbitofrontal Volume" = "medial orbito frontal", "Middle Temporal Volume" = "middle temporal", "Paracentral Volume" = "para central", "Parahippocampal Volume" = "parahippocampal", "Pars Opercularis Volume" = "pars opercularis", "Pars Orbitalis Volume" = "pars orbitalis", "Pars Triangularis Volume" = "pars triangularis", "Pericalcarine Volume" = "pericalcarine", "Postcentral Volume" = "post central", "Posterior Cingulate Volume" = "posterior cingulate", "Precentral Volume" = "pre central", "Precuneus Volume" = "precuneus", "Rostral Anterior Cingulate Volume" = "rostral anterior cingulate", "Rostral Middle Frontal Volume" = "rostral middle frontal", "Superior Frontal Volume" = "superior frontal", "Superior Parietal Volume" = "superior parietal", "Superior Temporal Volume" = "superior temporal", "Supramarginal Volume" = "supramarginal", "Temporal Pole Volume" = "temporal pole", "Transverse Temporal Volume" = "transverse temporal")
      rownames(volumes) <- volumes$Region
  
       #select out subcortical measures and rename to match ggseg conventions
      subcort <- results_adjusted[c("Amygdala Volume", "Caudate Volume", "Hippocampus Volume", "Lateral Ventricles Volume", "Pallidum Volume", "Putamen Volume", "Thalamus Volume", "Ventral DC Volume"),]
   
      subcort$Region <- dplyr::recode(subcort$Region, "Amygdala Volume" = "amygdala", "Caudate Volume" = "caudate",  "Hippocampus Volume" = "hippocampus", "Lateral Ventricles Volume" = "lateral ventricle", "Pallidum Volume" = "pallidum", "Putamen Volume" = "putamen", "Thalamus Volume" = "thalamus proper", "Ventral DC Volume" = "ventral DC")
      rownames(subcort) <- subcort$Region

      
      
      
      ####### Graph Age model beta values ########
      
      theme <- theme(axis.text.x=element_blank(), 
                     axis.title.x=element_blank(), 
                     legend.text=element_text(size=10),
                     legend.title=element_text(size=12,face="italic"),
                     legend.position="right", 
                     plot.title=element_text(size=10,hjust=0.5) )
      
      scale1 <- scale_fill_gradient(low="navyblue", high="gold",na.value="grey", limits=c(-0.75,0.08)) 
      scale2 <- scale_fill_gradient2(low=("#f46d43"), mid="white",  high=("#74add1"), midpoint=0,na.value="grey", limits=c(-1,1))
      
      
      
      #Graphs the age linear model betas
      thick_fig <- ggseg(.data=data.frame(area=row.names(thicknesses), Beta=(thicknesses$strength_beta)), mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3, show.legend=F) +  
        labs(title="Thicknesses") + 
        scale1 + 
        theme
      
      
      vol_fig <- ggseg(.data=data.frame(area=row.names(volumes), Beta=(volumes$strength_beta)), mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3, show.legend=F) +  
        labs(title="Volumes") + 
        scale1 + 
        theme
      
      
      aseg_fig <- ggseg(.data=data.frame(area=row.names(subcort), Beta=(subcort$strength_beta)), atlas="aseg", mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3) +  
        labs(title="") + 
        scale1 + 
        theme 
      
      
      age_lm_plots <- plot_grid(thick_fig, vol_fig, ncol=1, nrow=2, align="H")
      age_lm_plots <- plot_grid(age_lm_plots, aseg_fig , ncol=1, nrow=2, align="H")
      #age_lm_plots
      
      
      
      ########### Graphs the slope age linear model ########
      thick_fig <- ggseg(.data=data.frame(area=row.names(thicknesses), Beta=(thicknesses$pattern_beta)), mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3, show.legend=F) +  
        labs(title="Thicknesses") + 
        scale2 + 
        theme
      
      
      vol_fig <- ggseg(.data=data.frame(area=row.names(volumes), Beta=(volumes$pattern_beta)), mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3, show.legend=F) +  
        labs(title="Volumes") + 
        scale2 + 
        theme
      
      
      aseg_fig <- ggseg(.data=data.frame(area=row.names(subcort), Beta=(subcort$pattern_beta)), atlas="aseg", mapping=aes(fill=Beta), hemisphere="left", colour="black", size=.3) +  
        labs(title="") + 
        scale2 + 
        theme
      
      
      slope_lm_plots <- plot_grid(thick_fig, vol_fig , ncol=1, nrow=2, align="H")
      slope_lm_plots <- plot_grid(slope_lm_plots, aseg_fig , ncol=1, nrow=2, align="H")
      #slope_lm_plots
      
      
      
      ###### Combine Age and Slope beta weight graphs #########
      full_lm_plot <- plot_grid(age_lm_plots, slope_lm_plots , labels=c("Strength", "Pattern"),label_size=16,  ncol=2, nrow=1, align="H")
      
      return(full_lm_plot)
      

  }


  output$region_plot <- renderPlot({
    
   
  #do the calculation once to set the y axis - then repeat with the specific choices made
  plotFrame <- fit_data(normals,input$ROI_name, 0.7)
  #Set y-axis (doing this early and then rerunning for subset options so y axis doesn't change)
  y_max <- (plotFrame[order(plotFrame$fit + 3*plotFrame$sd, decreasing=T),][1,]$fit + 3*plotFrame[order(plotFrame$fit + 3*plotFrame$sd, decreasing=T),][1,]$sd) #normals max + 2.5SD (at the max)
  y_min <- (plotFrame[order(plotFrame$fit - 3*plotFrame$sd),][1,]$fit - 3*plotFrame[order(plotFrame$fit - 3*plotFrame$sd),][1,]$sd) #lowest normals data - 2.5 SD (at the min) 
  region_customY <- scale_y_continuous(limits=c(y_min,y_max))
  
  
  #set x-axis 
  x_min <- n_min - 1
  x_max <- n_max
  region_customX <- scale_x_continuous( breaks=(seq(from=(10*floor(x_min/10)), to=10*ceiling(x_max/10), by=10)),  limits=c(x_min, x_max))
  
  
  
  #select out amyloid positives unless selected
  if(!input$pcad){normals <- normals[normals$AD_status=="Amyloid-",]}
  
  #choose subset then repeat calculation
  if(input$sex==2){normals <- normals[normals$Sex=="M",]}
  if(input$sex==3){normals <- normals[normals$Sex=="F",]}
  if(input$apoe4==2){normals <- normals[normals$apoe4==0,]}
  if(input$apoe4==3){normals <- normals[normals$apoe4==1,]}
  
  
  plotFrame <- fit_data(normals, input$ROI_name, input$span)
  
  slope <- diff(plotFrame$fit, differences=1)
  slope <- (slope) / plotFrame$fit[-length(plotFrame$fit)] * 100  #make it percent change
  plotFrame$slope <-c(NA, slope)
    
  #### Graph Making
  
  #ggplot2 options
  options <- theme(axis.text.x=element_text(size=18),axis.text.y=element_text(size=18),
                   axis.title.x=element_text(size=18,face="bold"),
                   axis.title.y=element_text(size=18,angle=90,face="bold"),
                   legend.text=element_text(size=18),
                   legend.title=element_blank(),legend.position="right",
                   plot.title=element_text(size=18,face="bold",hjust=0.5),legend.title.align=0.5,
                   plot.subtitle=element_text(size=16))
  
  #Make Plot elements from normals data
  region_plot <- ggplot(plotFrame,aes(x=Age,y=fit))
  region_smoothLine <- geom_line(aes(x=Age,y=fit,colour="Normals"),size=1.25)
  region_smoothRibbon2 <- geom_ribbon(aes(x=Age,ymax=fit+2*sd,ymin=fit-2*sd, fill="2 Std Dev"),alpha=0.1)
  region_smoothRibbon <- geom_ribbon(aes(x=Age,ymax=fit+sd,ymin=fit-sd, fill="1 Std Dev"),alpha=0.5)
  
  #Make Aesthetic Plot elements
  region_xlabel <- xlab("Age") #x label, doesn't change
  if (grepl("thickness", input$ROI_name, ignore.case=T)) {
    region_ylabel <- ylab(expression(bold(paste('Thickness (mm)'))))
    is_volume <- FALSE
  } else {
    region_ylabel <- ylab(expression(bold(paste('Normalized Volume  ',(mm^{3})))))
    is_volume <- TRUE
  }
  
  #set title
  region_title <- ggtitle(input$ROI_name, subtitle=NULL)
  
  
  
  
  
  #set legend labels - do not mess with, reorders labels without clear cause  
  legend  <- scale_colour_manual(name="", labels=c("Average"),  
                                 values=c("1 Std Dev"="#F8766D", "Normals"="red3", "2 Std Dev"="lightblue"))  
  
  
  #add dotplot of the individual subjects on top of the fit line
  if(input$pcad){
  originals <- geom_point(data=normals, aes(x=normals$Age, y=normals[,input$ROI_name], shape=factor(AD_status) )) 
  originals_shape <-  scale_shape_manual(values = c(19, 2))
  #adds scatter plot of original oasis data overtop the fit line    
  }else{
  originals <- geom_point(data=normals, aes(x=normals$Age, y=normals[,input$ROI_name] )) #adds scatter plot of original oasis data overtop the fit line 
  originals_shape <-  scale_shape_manual(values = c(19))
  }

  
  
  Fig <- region_plot + theme_bw() + options  + region_smoothRibbon2 + region_smoothRibbon  + region_smoothLine + region_xlabel + region_ylabel + region_title
  
  #if adding patient data
  if(!is.na(input$region_val) & !is.na(input$Age) & (!is.na(input$ICV))){
    #Set up if patient is within the age range covered by super normal cohort
    if( (input$Age >=min(normals$Age, na.rm=TRUE)) & (input$Age <=max(normals$Age, na.rm=TRUE)) ){
      ages_within_normals <- TRUE  
      }else{  ages_within_normals <- FALSE  }
    
    #Correct for head volume size
    if(is_volume & !(input$ROI_name %in% c("Intracranial Volume", "White Matter Hyperintensities Volume", "Non-WM Hyperintensities Volume"))){
      region_val_corrected <- data.frame( input$region_val - as.numeric(ICV_correction_coeff[input$ROI_name])*(input$ICV -     mean(normals$`Intracranial Volume`) )) #normalizes the patient data based on ICV
    }else{region_val_corrected <- as.numeric(input$region_val) }
    
    
    #Calculate z scores and associates percentiles for each of patient's MR data points
    region_z_score=matrix()
    region_percentile=matrix()
    
    if (ages_within_normals==TRUE){
      smoothFit <- loess( as.matrix(normals[,input$ROI_name]) ~ normals$Age, degree=1, span=input$span) #locally fits a polynomial surface
      stand_dev2 <- matrix()
      for (i in 1:length(n_min:n_max)) {
        stand_dev2[i] <- sd(as.matrix(normals[normals$Age > seq(n_min,n_max, by=1)[i]-1 & normals$Age < seq(n_min,n_max, by=1)[i]+1,][,input$ROI_name]))
      }
      sd_fit <- loess(stand_dev2 ~ c(n_min:n_max), degree=1, span=input$span)
      region_z_score=((region_val_corrected - predict(smoothFit, newdata=input$Age)) / 
                          predict(sd_fit, newdata=input$Age)) #Calculated z score of patient's data
      
      region_percentile <- round( pnorm(as.numeric(region_z_score))*100, digits=0 ) #Converts z score into a percentile
      
      if (as.numeric(region_percentile)<1) { region_percentile=", 1st Percentile" } #defines percentile if below limit
      
      else if (as.numeric(region_percentile)>99) { region_percentile=", 99th Percentile" } #defines percentile if above limit
      
      else if ((as.numeric(region_percentile)>=1) && (as.numeric(region_percentile)<=99)){
        region_percentile=paste(", ", toOrdinal(region_percentile)," Percentile", sep="") #add a comma to the version being used in the display
      }
      
    }else {
      region_percentile=""  #leaves percentile blank if patient is outside age range of normal group
    }
    
    #make ggplot element
    region_scatterNorm <- geom_point(aes(x=input$Age,y=as.numeric(region_val_corrected),group="Patient", colour="Patient"),size=4.0) #graphs patient's datapoints
    
    #Add text on patient's data points and calculated percentiles
    #Set up the label
    
    display_label=paste(signif(region_val_corrected, digits=4), t(region_percentile), sep="")
    
    #Use ggrepel to display the label so it doesn't get cut off
    region_volume_text <- geom_text_repel(data=data.frame(input$Age, as.numeric(region_val_corrected)), mapping=aes( x=input$Age, y=as.numeric(region_val_corrected), label=display_label), fontface="bold", size=6, point.padding=unit(.5, 'lines'), box.padding=unit(0.5, 'lines'), force=5 )
    
    #Set y max
    
    if (y_max < region_val_corrected + 0.5*mean(plotFrame$sd, na.rm=TRUE)) {
      y_max <- as.numeric(region_val_corrected + 0.5*mean(plotFrame$sd, na.rm=TRUE)) #use patient's y max + 0.5 SD as graph max
    }
    
    #Set y min
    if (y_min > region_val_corrected - 0.5*mean(plotFrame$sd, na.rm=TRUE)) {
      y_min <- as.numeric(region_val_corrected - 0.5*mean(plotFrame$sd, na.rm=TRUE)) #use patient's y max + 0.5 SD as graph max
    }
    
  
  region_customY <- scale_y_continuous(limits=c(y_min,y_max))
      
  
       #set x-axis values
       if (input$Age > n_max) {x_max <- max(input$Age, na.rm=TRUE) }
       if (input$Age < n_min){x_min <- min(input$Age, na.rm=TRUE) }
       region_customX <- scale_x_continuous( breaks=(seq(from=(10*floor(x_min/10)), to=10*ceiling(x_max/10), by=10)), limits=c(x_min, x_max))

     
       #set legend labels - do not mess with, reorders labels without clear cause
       legend  <- scale_colour_manual(name="", labels=c("Average", "Patient"),
                                      values=c("1 Std Dev"="#F8766D", "Normals"="red3", "Patient"="#673dbd", "2 Std Dev"="lightblue"))
     
       Fig <- Fig + region_scatterNorm + region_volume_text
    
  } #end of adding patient data
  
  
  #Combine the different elements of the plot
  Fig <- Fig + region_customY + region_customX + legend
  
  if (input$individual_points) {  Fig <- Fig + originals + originals_shape } 
  
  #print(Fig)  
  y_max <- 1
  y_min <- -1
  if ((min(plotFrame$slope, na.rm = T)-.2) < y_min){ y_min <- (min(plotFrame$slope, na.rm = T)-.2)  }
  if ((max(plotFrame$slope, na.rm = T)+.2) > y_max){ y_max <- (max(plotFrame$slope, na.rm = T)+.2)  }
  region_customY <- scale_y_continuous(limits=c(y_min,y_max))
  
  slope_plot <- ggplot(plotFrame, aes(x=Age,y=slope)) +
  theme_bw() + 
  options +
  geom_point(aes(color="Points")) + 
  xlab("Age") +  
  ylab(expression(bold('Rate of Change (%)'))) + 
  region_customX + 
  region_customY +
  scale_colour_manual(name="", labels=c("Points"="Rate"),  values = c("Points"="black"))  
  #print(slope_plot)

  plot_grid(Fig, slope_plot, ncol = 1, nrow = 2, align = "v")

  }, width=1100, height=1000)

  
  output$table <- renderTable({
    #select out amyloid positives unless selected
    if(!input$pcad){normals <- normals[normals$AD_status=="Amyloid-",]}
    #select correct subset
    if(input$sex==2){normals <- normals[normals$Sex=="M",]}
    if(input$sex==3){normals <- normals[normals$Sex=="F",]}
    if(input$apoe4==2){normals <- normals[normals$apoe4==0,]}
    if(input$apoe4==3){normals <- normals[normals$apoe4==1,]}
    #make table
    table <- region_map_data(normals)
    table
    }) 
  
  output$ggseg_plot <- renderPlot({ 
    #select out amyloid positives unless selected
    if(!input$pcad){normals <- normals[normals$AD_status=="Amyloid-",]}
    #select correct subset
    if(input$sex==2){normals <- normals[normals$Sex=="M",]}
    if(input$sex==3){normals <- normals[normals$Sex=="F",]}
    if(input$apoe4==2){normals <- normals[normals$apoe4==0,]}
    if(input$apoe4==3){normals <- normals[normals$apoe4==1,]}
    #make graph
    plot <- ggseg_graph(region_map_data(normals))
    plot
  }, width=900, height=800)
})

shinyApp(ui=ui, server=server,  options=list(height=900, width=1600))

  
```
